Chain separation distribution in protein contact networks

### 2020-08-06 paper/main.tex
nora changes

\begin{abstract}
The distribution of chain separations of interacting amino acids in proteins roughly follows a power law. Here we show, analytically and in simulations, that a geometrical stochastic model of folding chains can explain this behaviour. 
\end{abstract}
\maketitle

\section*{Introduction}
Proteins, the molecular machines of every living organism perform vital tasks required for life to persist, ranging from transport (e.g. hemoglobin), signal transduction (e.g. rhodopsin), immune responses (e.g. antibodies), and hormonal regulation (e.g. insulin)~\cite{something}. All natural proteins are made up of 20 different amino acids which dictate the three dimensional conformations the proteins adopt in order to function~\cite{stuff}. One way of looking at the functional forms of proteins and classifying their structural properties is by looking at protein residue or contact networks~\cite{Vendruscolo2002,DiPaola2013,Estrada2011}. Protein Contact Networks (PCN), have shown promise in understanding protein folding patterns as well as revealing allosteric communication pathways~\cite{https://pubs.acs.org/doi/10.1021/acs.jcim.9b00320, and others}. 
From a physicist perspective it is interesting to understand the emergent behaviour in protein structure from PCNs, by investigating common network measures and their behaviour for PCNs. For example Bartoli et al.~\cite{bartoli2008effect} build a model in which they assume that proteins have a sequence separation distribution that follows $\frac{1}{\ell}$, where $\ell$ is the separation of two connected amino acids along the backbone chain as the simplest implementation of a connection probability, that decreases with the distance along the sequence. However, they do not provide a quantitative explanation for this assumption.

@@

Based on such considerations, the network formation process, repeatedly \red{(this may not be the adverb i'm looking for, but subsequently also doesnt quite work)} connects random pairs of units, without causing any intersections along the chain ~\cite{molkenthin2016scaling}. Such a model can now be used to investigate the emergent behaviour of sequence separation distributions as observed by Bartoli et al.~\cite{bartoli2008effect}. To this end, we analyze the sequence separation distributions that emerge from the geometric model, where the \emph{sequence separation} $\ell$ of a link is defined as the distance of the two end units along the original chain.
This is then compared to the sequence separation distribution extracted from protein contact networks generated from protein data bank data.

\red{We find that, despite the highly simplified model, the generated sequence distance distribution is in agreement with the one observed in networks generated from PDB data and roughly consistent with the heuristic $1/\ell$ finding. The analytical approximation provides a correction term on top of the $1/\ell$ heuristic.}

@@

\section*{Analytic approximation}
Here we want to analytically approximate the sequence separation distribution for the simplified 2D model, that is the separation along the backbone chain of two nodes connected by a link, as illustrated in Fig~\ref{fig:schematic}~a).

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/FigFig1.pdf}
    \caption{a) The sequence separation distribution is defined as the separation along the chain (blue) of two connected units (red). b) The solid blue link prohibits some connections from node $A$ (dotted red), while others are still permitted (dotted blue). Counting all newly excluded links, we find that a link of length $k$ prohibits $2(l-1)$ links of length $l$ up to a maximum of $k$ and $2(k-1)$ links of length $k$ or more.}
    \label{fig:schematic}
\end{figure}

Let us start by introducing the auxiliary variable $F(s)$, defined to be the number of possible links with a sequence separation of $s$. Before any links are added all sequence separations are equally likely, as there are $N$ possibilities of making a link of each separation $s$ with $2\leq s < N/2$.

As we add more links, not only are links taken out of this pool, because they have already been realized, each existing link can also geometrically prohibit other connections. Adding a link of length $s_1$ prohibits $2(s-1)$ links of length $s<s_1$ and $2(s_1-1)$ links of length $s>s_1$ (see Fig.~\ref{fig:schematic}~b). To find the expected number of links still available in subsequent stops, we average over all possible sequence separations $s_1$ of the first link added.

@@

The probability distribution of the realized sequence separations of all added links is then given by the average over the available pools at each link addition step, leading to
\begin{align}
    P(s)&=\frac{1}{N-3}\sum_{k=0}^{N-3} P_k(s) \nonumber \\
    &= \frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C_{i-1}} \right),
\end{align}
which makes use of Eq.~\ref{eq.Pk}. The only unknown in this expression is $C_k$, which we can not compute exactly. However, since $C_k$ monotonically decreases with $k$, $N>C_k>C_{N-2}$ always holds and we can obtain an upper and lower bound by replacing all $C_k$ in the expression with either $N$ (underestimates all probabilities) or $C_{N-2}$ (overestimates all probabilities). Each time we get an expression of the form 
\begin{align}
     P(s)&\approx\frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)\nonumber \\
     &=\frac{N}{N-3} \sum_{k=0}^{N-3}\frac{1}{C_k}\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)^k \nonumber \\
     &\approx \frac{A}{f(s)}
\end{align}
for $1<s<\frac{N}{2}$,
where $A$ is an unknown constant, containing an approximation of the various normalization constants $C_i$, as well as the direct effects of the chain length. 
Since for long chains, $f(s)$ is largely dominated by the linear term, the distribution is well approximated by a power law with an exponent of $-1$. We have made use of the geometric series to approximately find a closed solution.

@@

\section*{Data and simulations}
 For large values of the chain length $N$ we thus find an approximate power law probability distribution for the sequence separation with an exponent of $-1+\alpha$. Here we use simulations, as introduced in \cite{molkenthin2020self} and networks extracted from measured protein data from the PDB \cite{PDB} to compare this behaviour in realistic 3D settings and show that the basic principle persists, indicating that much of the sequence distance separation distribution can be explained by simple geometric constraints.
 \begin{figure}[h]
        \centering
	\includegraphics[width=0.45\textwidth]{figures/pdb_statistics.pdf}
        \caption{Statistics on all PDB entries based on the number of amino acid residues, with most entries found in the interval of 100-400 residues as of May 26th 2020.}
        \label{fig:pdb_stats}
\end{figure}

Protein chain lengths in the PDB vary from small fragments of less than 10 amino acids to large agglomerates of over 2000 residues. The most frequently occurring chain lengths, however are between 100 and 400 residues as shown in Fig.~\ref{fig:pdb_stats}. For the analysis of our theoretical results, we have thus chosen ............ structures of around 200, ........ structures of around 300 and .......... structures of around 400 residues. Those have been converted to PCN's

\begin{figure*}[t]
        \centering
	\includegraphics[width=\textwidth]{paper/figures/resultsfig_old.pdf}
        \caption{Sequence distance distributions for real and simulated folds for a) Chain lengths around 200 b) chain lengths around 300 c) chain lengths around 400.
        }
        \label{fig:sdd}
\end{figure*}
We observe that, while the general shape seems to match, in real proteins sequence separations of 2 to 4 as well as 20 to 50 amino acids are slightly over-represented (see Fig.\ref{fig:sdd}) \red{I have a vague memory of figures with more data where the hump was less pronounced? Depending on this rephrase the above.}.

The over-representation of very short sequence separations is an artefact of the different network construction for simulated networks and PRNs. While in the simulation all links are equally long and links are made until no longer possible, the parameter, determining the connectivity in PRNs is the cutoff distance $d_c$, which is chosen such that the number of links in the PRN is comparable to those in a simulated network of the same $N$. This leads to a larger range of bond angles randomly resulting in connections for small $s$ and thus an over-representation of those links.

\section*{Conclusion}
We have demonstrated in analytical calculations and direct simulations how a random linking models with geometric constraints, such as the ones introduced in \cite{molkenthin2016scaling, molkenthin2020self} generate sequence distance distributions, that resemble a power-law, similar to the one found in folded proteins. This feature of protein network structure has been used to generate protein-like structures in \cite{bartoli2008effect}.

\red{I remember we agreed on just deleting all the stuff about idp's beause the bump wasn't very pronounced once we used more data and thus the whole narrative of helix lengths and stuff wasn't particularly convincing. Additionally this means less work and thus sooner publication, so we were actually quite happy about it. But then we need to come up with something to write in the conclusion, so far it's a bit short.}

>>>>

\begin{abstract}
The mechanisms by which the folded structures of proteins are determined has long been one of the key mysteries of Biophysics. Here we approach the problem from a network ensemble perspective, analyzing the relative frequencies with which residues at different distances along the chain come into contact. We define and analyze the residue distance distribution for measured protein structures and structures resulting from a geometrically constrained random interaction model and find that the distributions are compatible. We furthermore derive an analytic approximation for the two dimensional analog of the model and find this approximation to also describe the distribution. We therefore propose the conclusion that the approximate power-law behaviour of the residue distance distribution is largely caused by geometric constraints of chains rather than specifics of protein sequences.
%The distribution of chain separations of interacting amino acids in proteins roughly follows a power law. Here we show, analytically and in simulations, that a geometrical stochastic model of folding chains can explain this behaviour. 
\end{abstract}
\maketitle

\section*{Introduction}
Proteins, the molecular machines of every living organism perform vital tasks required for life to persist, ranging from transport (e.g. hemoglobin), signal transduction (e.g. rhodopsin), immune responses (e.g. antibodies), and hormonal regulation (e.g. insulin)~\cite{something}. All natural proteins are made up of 20 different amino acids which dictate the three dimensional conformations the proteins adopt in order to function~\cite{stuff}. One way of looking at the functional forms of proteins and classifying their structural properties is by looking at protein residue or contact networks~\cite{Vendruscolo2002,DiPaola2013,Estrada2011}. Protein Contact Networks (PCN), have shown promise in understanding protein folding patterns as well as revealing allosteric communication pathways~\cite{https://pubs.acs.org/doi/10.1021/acs.jcim.9b00320, and others}. 
From a physicist perspective it is interesting to understand the emergent behaviour in protein structure from PCNs, by investigating common network measures and their behaviour for PCNs. For example Bartoli et al.~\cite{bartoli2008effect} build a model in which they assume that proteins have a residue distance distribution that follows $\frac{1}{\ell}$, where $\ell$ is the separation of two connected amino acids along the backbone chain as the simplest implementation of a connection probability, that decreases with the distance along the sequence. However, they do not provide a quantitative explanation for this assumption.

@@

Based on such considerations, the network formation process, repeatedly connects random pairs of units, without causing any intersections along the chain ~\cite{molkenthin2016scaling}. Such a model can now be used to investigate the emergent behaviour of residue distance distributions as used heuristically by Bartoli et al.~\cite{bartoli2008effect}. To this end, we define and analyze the residue distance distributions $P(s)$ that emerge from the geometric model, where the \emph{residue distance} $s$ of a link is defined as the distance of the two end units along the original chain.
This is then compared to the residue distance distribution extracted from protein contact networks generated from protein data bank data.

We find that, despite the highly simplified model, the generated residue distance distribution is in agreement with the one observed in networks generated from PDB data and roughly consistent with the heuristic $1/s$ finding. The analytical approximation provides a correction term on top of the $1/s$ heuristic.

@@

\section*{Analytic approximation}
Here we want to analytically approximate the residue distance distribution for the simplified 2D model, that is the separation along the backbone chain of two nodes connected by a link, as illustrated in Fig~\ref{fig:schematic}~a).

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/Fig1/Fig1.pdf}
    \caption{a) The residue distance distribution is defined as the separation along the chain (blue) of two connected units (red). b) The solid blue link prohibits some connections from node $A$ (dotted red), while others are still permitted (dotted blue). Counting all newly excluded links, we find that a link of length $k$ prohibits $2(l-1)$ links of length $l$ up to a maximum of $k$ and $2(k-1)$ links of length $k$ or more.}
    \label{fig:schematic}
\end{figure}

Let us start by introducing the auxiliary variable $F(s)$, defined to be the number of possible links with a residue distance of $s$. Before any links are added all residue distances are equally likely, as there are $N$ possibilities of making a link of each separation $s$ with $2\leq s < N/2$.

As we add more links, not only are links taken out of this pool, because they have already been realized, each existing link can also geometrically prohibit other connections. Adding a link of length $s_1$ prohibits $2(s-1)$ links of length $s<s_1$ and $2(s_1-1)$ links of length $s>s_1$ (see Fig.~\ref{fig:schematic}~b). To find the expected number of links still available in subsequent stops, we average over all possible residue distances $s_1$ of the first link added.

@@

The probability distribution of the realized residue distances of all added links is then given by the average over the available pools at each link addition step, leading to
\begin{align}
    P(s)&=\frac{1}{N-3}\sum_{k=0}^{N-3} P_k(s) \nonumber \\
    &= \frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C_{i-1}} \right),
\end{align}
which makes use of Eq.~\ref{eq.Pk}. The only unknown in this expression is $C_k$, which we can not compute exactly. However, since $C_k$ monotonically decreases with $k$, $N>C_k>C_{N-2}$ always holds and we can obtain an upper and lower bound by replacing all $C_k$ in the expression with either $N$ (underestimates all probabilities) or $C_{N-2}$ (overestimates all probabilities). Each time we get an expression of the form 
\begin{align}
     P(s)&\approx\frac{N}{N-3}\sum_{k=0}^{N-3} \frac{1}{C_k}\prod_{i}^k\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)\nonumber \\
     &=\frac{N}{N-3} \sum_{k=0}^{N-3}\frac{1}{C_k}\left(1-\frac{f(s)}{(\frac{N}{2}-2)C} \right)^k \nonumber \\
     &\approx \frac{A}{f(s)}
\end{align}
for $1<s<\frac{N}{2}$,
where $A$ is an unknown constant, containing an approximation of the various normalization constants $C_i$, as well as the direct effects of the chain length. 
Since for long chains, $f(s)$ is largely dominated by the linear term, the distribution is well approximated by a power law with an exponent of $-1$. We have made use of the geometric series to approximately find a closed solution.

@@

\section*{Data and simulations}
For large values of the chain length $N$ we thus find an approximate power law probability distribution for the residue distance with an exponent of $-1+\alpha$. Here we use simulations, as introduced in~\cite{molkenthin2020self} and networks extracted from measured protein data from the PDB \cite{PDB} to compare this behaviour in realistic 3D settings and show that the basic principle persists, indicating that much of the residue distance distribution can be explained by simple geometric constraints.
 \begin{figure}[h]
        \centering
	\includegraphics[width=0.45\textwidth]{paper/figures/Fig2/pdb_statistics.pdf}
        \caption{Statistics on all PDB entries based on the number of amino acid residues, with most entries found in the interval of 100-400 residues as of May 26th 2020.}
        \label{fig:pdb_stats}
\end{figure}

Protein chain lengths in the PDB vary from small fragments of less than 10 amino acids to large agglomerates of over 2000 residues. The most frequently occurring chain lengths, however are between 100 and 400 residues as shown in Fig.~\ref{fig:pdb_stats}. \red{For our analysis, we have thus chosen ............ structures of around 200, ........ structures of around 300 and .......... structures of around 400 residues from the PDB.} Those have been converted to PCN's based on the measured locations of their residues to analyze the distribution of residue distances between connected amino acids. The resulting distribution is compared to those generated in simulations in~\cite{molkenthin2020self} as well as the above derived theoretical expression. The latter, however, is scaled with a factor to account for the different dimensionality, as well as approximations made in the derivation.

The results are shown logarithmically scaled in Fig.\ref{fig:sdd}. \red{For all three lengths, the distributions show an approximate power law decay with slopes of $...\pm ...$, which is consistent with both, the simulated results ans the theoretical approximation.}

\begin{figure}[t]
        \centering
	\includegraphics[width=0.45\textwidth]{paper/figures/Fig3/Fig3.pdf}
        \caption{residue distance distributions for real and simulated folds for a) Chain lengths around 200 b) chain lengths around 300 c) chain lengths around 400. \red{Depending on what this looks like when we have all the data, perhaps adding error bars could be nice. Also, since i'm mentioning the overrepresentation of separations 2 to 4, maybe those should be shown. If it fits better we can take them out for the fits and stuff.}
        }
        \label{fig:sdd}
\end{figure}

We observe that, while the general shape seems to match, in real proteins residue distances of 2 to 4 amino acids are slightly over-represented (see Fig.\ref{fig:sdd}).
The over-representation of very short residue distances is an artefact of the different network construction for simulated (and theoretically approximated) networks and PCNs. While in the simulation all links are equally long and links are made until no longer possible, the parameter, determining the connectivity in PCNs is the cutoff distance $d_c$, which is chosen such that the number of links in the PCN is comparable to those in a simulated network of the same $N$. This leads to a larger range of bond angles randomly resulting in connections for small $s$ and thus an over-representation of those links.

\section*{Conclusion}
The tertiary structures of folded proteins have long been one of the most important mysteries of Biophysics. While for individual structures, detailed molecular dynamics or statistical models are essential in approaching these questions, general statements for the ensemble of folded proteins can be made with simple models.

Here we have introduced and analyzed the residue distance $s$ and its probability distribution $P(s)$ in measured and simulated protein structures. For the equivalent model in two dimensions we have used a mean-field approach to derive an analytic approximation for the residue distance distribution, which appears to agree with the simulated and measured distributions.
We have thus demonstrated in analytical calculations as well as direct simulations how a random linking model with geometric constraints, such as the ones introduced in \cite{molkenthin2016scaling, molkenthin2020self} generates a residue distance distribution, that resembles a power-law, similar to the one found in folded proteins. This feature of protein network structure has previously been used in the heuristic model to generate protein-like structures in \cite{bartoli2008effect}.

Gaining a better understanding of the ensemble of folded protein structures can help guide the way to a better understanding of the constraints within which structures may occur. Together with an understanding of secondary structure principles, such as those in
\cite{Danielsson2010, Molkenthin2011}, this can help to narrow down the complex energy landscapes and find paths through them more effectively in the future.
